Doublet splitting of fusion alpha-particle driven ion cyclotron emission

### 2022-07-28 manuscript.tex
careful re-reading improvements

Observations of ion cyclotron emission (ICE) have been reported from
magnetically confined plasmas first in TFTR\cite{Cauffman1995,Dendy1995}
and JET\cite{Cottrell1988,Cottrell1993,Dendy1995}, followed by other tokamaks
PDX\cite{Heidbrink1994}, JT-60-U\cite{Ichimura2008}, DIII-D\cite{Heidbrink2011},
ASDEX\cite{DInca2014}, KSTAR\cite{Chapman2017,Chapman2018},
ASDEX-Upgrade\cite{Ochoukov2018},
TUMAN-3M\cite{Askinazi2018}, EAST\cite{Liu2020}. Stellarators\cite{Saito2013} and
FRCs\cite{Nicks2020} have also exhibited this phenomenon. Energetic ion
populations that give rise to ICE have their origins in fusion
reactions in DT plasmas\cite{Cottrell2000, Cauffman1995},
ICRH\cite{Cottrell2000}, and neutral beam injection\cite{Ochoukov2018}. ICE
arising from fusion $\alpha$-particles is a proposed route to further
the understanding of burning plasmas in ITER\cite{Dendy2015,McClements2015}.

The foremost explanation of the mechanism responsible for ICE is the
magnetoacoustic cyclotron instability (MCI)\cite{Belikov1976}, which arises from
the free energy present in energetic ions with inverted distributions in
velocity space. Theoretical understanding of ICE is based in analytical
solutions to the linearised Maxwell-Vlasov system of equations\cite{Stix} where
the energetic ion species was modelled as ring-beam for which $f(v_\parallel,
v_\perp) \propto \exp(-(v_\parallel-u_\parallel)^2/ v_{th}^2)\delta(v_\perp -
u_\bot)$ \cite{Dendy1994, McClements1996}. Here $u_\parallel$, $u_\perp$ and
$v_{th}$ are the beam velocity (the signed drift speed along the magnetic
field), ring speed (the particle speed perpendicular to the magnetic field)
and thermal velocity, respectively. Subsequent studies augmented this
understanding via nonlinear fully-kinetic\cite{Cook2013,Chapman2017} and
hybrid\cite{Carbajal2014,Reman2019} particle-in-cell (PIC) simulations in one
spatial dimension and three vector components for velocities and electromagnetic
field (i.e. 1D3V), and recently 2D3V hybrid-PIC simulations\cite{Carbajal2021}. These
computational approaches operate under similar simplifications as the analytical
treatments in that the plasma is initially spatially homogeneous and periodic
in the spatial domain.
Linearised fluid approaches that incorporate toroidal geometry and
inhomogeneous plasma and magnetic fields have shed light of the eigenmode
structure of compressional Alfv{\'e}n eigenmodes, which as with ICE are also fast
Alfv{\'e}n modes driven by energetic ions, in spherical
tokamaks\cite{Sharapov2014, Gorelenkov2016} and conventional aspect ratio
tokamaks \cite{Kolesnichenko2000, Fulop2000, Heidbrink2006}.

The effectiveness of analytical theory to reproduce the observed cyclotron
harmonic peaks in the fluctuations of the magnetic field was first explained
somewhat by 1D PIC simulations\cite{Cook2013}.
In these simulations the onset of nonlinearity
appeared to imprint the amplitudes of each of the harmonics into the signal as
they are at the end of the linear phase; each mode grows independently until 
growth is terminated for all modes when the fastest mode enters its nonlinear
phase meaning that both the amount of free energy available for instability has
decreased and that the unstable ion distribution function has been
perturbed into a more stable state. Analysis shows that only higher cyclotron
harmonics are destabilised\cite{McClements2015} to the MCI under these
conditions, however experimental
observations clearly indicate power at theoretically stable lower
harmonics\cite{Cottrell1988}. In a further step towards accurate recreation of
experimental observations, nonlinear phases of hybrid\cite{Carbajal2014} and
fully kinetic\cite{Cook2013,Chapman2017} PIC simulations went on to show that
wave-wave coupling was responsible for raising the amplitude of lower cyclotron
harmonics. Most recently, hybrid 2D simulations have begun to shed light on the
2D3V dynamics of ICE\cite{Carbajal2021}.

Finer detailed structure has been obvserved in ICE spectra, such as doublet
splitting of individual harmonics on JET\cite{Cottrell1993} and
DIII-D\cite{Thome2018} and even multiplet splitting on TUMAN-3M\cite{Askinazi2018},
ASDEX-Upgrade\cite{Ochoukov2019} and EAST\cite{Lui2020}.
This paper highlights one underlying kinetic mechinism that produces
doublet splitting due to Doppler shifted resonances and the
intricate landscape of the MCI growth rate on the fast Alfv{\'e}n dispersion
surface. In principle, splitting could be produced by grad-B or curvature
drifts\cite{Cottrell1993,Fulop1998} or potentially any mechanism that is capable
of adding a Doppler shift to the wave-particle resonance condition. Note that
this study focusses on edge-ICE, named for the region of plasma from which the signal
arises, as opposed to core-ICE\cite{Ochoukov2018, Liu2020}.

To better enable ICE as a diagnostic for confined fast ions, it is necessary to
further develop our understanding with a range of tools. In this work, I present
the growth rates in 2D wavenumber space of the MCI under the conditions of
outer mid-plane of the JET PTE\cite{Cottrell1993} Pulse No. 26148. These growth
rates are obtained from a dispersion relation solver,
LinearMaxwellVlasov.jl\cite{LinearMaxwellVlasov.jl}, written in the Julia programming
language\cite{Bezanson2017}. These results narrow the gap between analytical
theory and expensive HPC simulations and can be run on desktop computers
in a matter of minutes, further enabling research into this topic.

In section 2, the computational method is described for calculating
the linear growth rates of the MCI, which are presented for energetic ring-beam
alpha-particles with non-zero perpendicular thermal spread in section 3
where a comparison is ade to the observations on JET..
Results are summarised in section 4.

@@

Solutions to the linearised Maxwell-Vlasov system of equations are found in
parallel and perpendicular wavenumber space, $(k_\bot, k_\parallel)$, for a
ring-beam distribution of energetic minority
alpha-particles and Maxwellian electrons and background thermal majority
deuteron ions. The ring-beam distribution function, Eq. 1, is a separable
function of the beam in $v_\parallel$ given by Eq. 2 and the
ring in $v_\perp$ given by Eq. 3.
In the context of the fully kinetic Maxwell-Vlasov linear theory,
which rests upon the assumption of
an infinite homogeneous plasma and a straight uniform magnetic field,
this distribution function is parameterised  by the drift
velocity parallel to the magnetic field $u_\parallel$, the drift speed 
perpendicular to the magnetic
field $u_{\perp}$, and the isotropric thermal velocity $v_{th}$. The
thermal spread of the alpha particles' distribution function 
$v_{th}=v_{\parallel,th}=v_{\perp,th}=u_0/100$. The number density
ratio of alpha-particles to the majority fuel deuterons is $1.5\times10^{-4}$.

@@

Although curvature and grad-B drift terms would be present in fast
alpha-particle populations on JET and these drifts would alter the wave-particle
resonance condition, it is worth noting that they are both of
order $\rho^*$ and have been neglected in these calculations where $u_\bot \gg
\rho^* u_0$ is always the case.

Solutions to the linearised Maxwell-Vlasov system of equations are given by the
dispersion relation, Eq. \ref{eq:dettensor}, which is satisfied for sets of
$(\omega,\vec{k})$ given the distribution functions of all the species present
along the fast Alfv{\'e}n branch  on a grid of $(k_\parallel, k_\bot)$.
The hot dielectric tensor\cite{Stix}, $\vec{\vec{\epsilon}}$ given by Eq.
\ref{eq:dielectrictensor} is calculated for each species, $s$,

@@

Integrals over perpendicular velocity required for calculating the
dielectric tensor for Maxwellian distribution functions are given in terms of
modified Bessel functions of the first kind\cite{Stix} for both electrons and
deuterons i.e. when $u_\perp =0$. The same perpendicular integrals over the
alpha-particles' ring distribution given by Eq. 3 are calculated numerically via
Gauss-Kronrod\cite{Kronrod1965} quadrature using the GaussGK.jl\cite{QuadGK}
library. An absolute and relative tolerance of $0$ and $10^{-8}$, respectively,
is used to confirm convergence of the integral values.
Since all the parallel distribution functions for all species are Maxwellians,
the parallel integrals can be calculated from the well known plasma dispersion
function\cite{Fried1961}, which is laid out in the appendix.

Bessel function indices, $n$, in Eq. \ref{eq:dielectrictensor} are summed outwards
from 0 in both the positive and negative directions
until the relative change between the current and
previous iteration is less than a tolerance of $10^{-8}$. 
Root finding of Eq. \ref{eq:dettensor} in the complex frequency plane
for a given $(k_\bot, k_\parallel)$ is performed by an amalgamation of the
Nelder-Mead\cite{Nelder1965} optimisation method with the winding number method.
If the complex values at the nodes of the Nelder-Mead simplex, which in this
case is just a triangle, have a winding number
of zero and therefore the simplex does not contain a root then the optimisation
procedure follows the Nelder-Mead method, otherwise the simplex bifurcates.
The child simplex that contains the root is used for the next
iteration. The optimisation iteration procedure finishes successfully when
a simplex that encircles a root is small enough such that its extent in real
(imaginary) frequency is less than $10^{-4}\, \Omega_i$ $(10^{-5}\, \Omega_i)$
, where $\Omega_i$ is the cyclotron frequency of both
the alpha-particles and deuterons since they share a charge-mass ratio.
This approach is found to be adequately robust and
faster than either the Nelder-Mead method or winding number approach used
separately. Each 512 by 512 scan in $(k_\perp, k_\parallel)$ takes between 5
and 6 minutes on a current laptop.

@@

The growth rates of unstable modes for the case where the pitch-angle cosine
value is zero ($u_\parallel = 0$ in Eq. 1) are plotted in Fig.
\ref{fig:2D_F12_zeropitch}(a) against the associated real frequency (only modes
with growth rates in excess of the solver tolerance are shown i.e. $\Im(\omega)
> 10^{-5} \Omega_i$). In this panel shading indicates parallel wavenumber in
units of $\Omega_i/V_A$. Fig. \ref{fig:2D_F12_zeropitch}(b) plots the same data
in the $(\omega,k_\parallel)$ plane where shading instead indicates growth
rates in units of $\Omega_i$.
The simplicity of the consecutive unstable cyclotron harmonics visible in
panel (a) betrays the complexity of the structure in the $(\omega,k_\parallel)$
plane. Integration over $k_\parallel$ projects the solutions onto the frequency
axis, which loses the structure and produces narrow harmonics. Figure
\ref{fig:2D_imag_zeropitch} gives $(k_\perp, k_\parallel)$ space view of the
same data as Fig. \ref{fig:2D_F12_zeropitch} as well as damped and
normal modes, except shading indicates the growth rates. Overlaid
contours indicate the real frequency in units of ion cyclotron harmonics. Only
narrow regions around the cyclotron harmonics are unstable. Overlaid straight
traces indicate propagation angle. No Doppler shift is present due to the lack
of parallel drift in the alpha-particle ring-beam. The growth rate increases
approximately linearly with cyclotron harmonic number, which is consistent with
the literature\cite{Dendy1994}. Together these figures serve to introduce a 2D
view (growth rates in $(k_\perp, k_\parallel)$ space) of traditionally 1D data
(growth rates against frequency). The importance of a non-zero $u_\parallel$ is
discussed next.

@@

Fig. \ref{fig:2D_F12_physicalpitch} plots solutions for the physically more
realistic pitch-angle
case ($u_\parallel = -0.646 u_0$) in the same manner as Fig.
\ref{fig:2D_F12_zeropitch} except there is an additional black trace in the
upper panel, which represents the measured JET ICE intensity signal; the continuous
trace from Fig. 2 in Ref. \cite{Cottrell1993}.
The observations of ICE intensity are plotted on a log scale scaled to match
the range of the linear MCI growth rates.
Direct comparison of log-scale
ICE observations and linear growth rates is justified to the extent that
nonlinear PIC simulations indicate that the linearly unstable modes rising
as $\exp(\gamma t)$, where $\gamma$ is a mode's growth rate,
cease to grow once the fastest growing mode becomes
nonlinear\cite{Cook2013}.
Therefore under the assumption that all modes start growing in unison from
identical seed amplitudes then the log of a mode's amplitude is
proportional to its growth rate.

@@

Doublet splitting can be seen in Fig. \ref{fig:2D_imag_physicalpitch}, which is the
counterpart figure to Fig. \ref{fig:2D_F12_physicalpitch} as Fig.
\ref{fig:2D_imag_zeropitch} is to Fig. \ref{fig:2D_F12_zeropitch},
at e.g. the $12^{th}$ harmonic  near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$, where $V_A$ is the Alfv{\'e}n speed.
Qualitatively, scattered points in Fig.
\ref{fig:2D_F12_physicalpitch}(a) resembles the experimental observations given
by the black trace in that the
signal below the $7^{th}$ cyclotron harmonics is lower and flatter with well
defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change
larger and the harmonics are less easily identifiable. These features cannot be
be recreated with a 1D calculation for these distributions functions
since they arise from contributions from a range of propagation angles.
Note that Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes
more clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is
characterised by narrow regions of instability near Doppler shifted cyclotron
harmonics that are stable when $k_\parallel=0$ and unstable at larger
negative parallel wavenumbers. The higher frequency behaviour in the region
$k_\bot \geq 7\Omega_i/V_A$ is dominated by broader bands of instability
near to Doppler shifted cyclotron harmonics, which have
an intricate structure in wavenumber space. Modes numbering 1, 2, 6, 7, 8, 12, and
13 are stable when propagating exactly perpendicular whilst obliquely
propagating modes remain unstable. The magnitude of the Doppler shift of
the growing modes can reach $\approx \Omega_i$ for large propagation angles at 
lower frequencies and also for
oblique angles at higher frequencies; see, for example, the mode
associated with the $15^{th}$ cyclotron harmonic near $k_\parallel \simeq
-\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 13 \Omega_i/V_A$, i.e. a propagation
angle of $\sim 93^\circ$. Furthermore, the breadth of the harmonics can be
attributed to the contributions from multiple propagation angles even though the
lower harmonics are largely stable for forward travelling modes and
propagation angles that are close to perpendicular.

Figure \ref{fig:2D_real_zeropitch}
shows the smoothed real frequency part of the solution for the case of
zero pitch angle cosine and is some normalised by
the high frequency expression for the fast Alfv{\'e}n wave set out in Eq.
25 of Ref. \cite{Dendy1994} (also Eq. 11 of Ref. \cite{McClements1996}),

@@

\noindent where $k^2 = k_\parallel^2 + k_\perp^2$. There is no discernable difference for the non-zero pitch angle cosine case.

@@

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig4.pdf}
    \caption{(colour online) As in Fig. \ref{fig:2D_imag_zeropitch}, except for
    a ring-beam of alpha particles with a more realistic pitch-angle value
    $u_{\parallel} = -0.646 u_0 \simeq -1.1 V_A$. The non-zero parallel drift
    speed causes up-down asymmetry; this figure for a pitch-angle cosine of the
    opposite sign, $u_{\parallel} = 0.646 u_0$, is identical except flipped
    about $k_\parallel=0$.}
    \label{fig:2D_imag_physicalpitch}
\end{figure}

@@

This paper presents the linear Maxwell-Vlasov dispersion relation of a plasma
consisting of thermal electrons, thermal deuteron fuel ions and a minority
energetic ring-beam of fusion born alpha-particles. These high resolution
$(k_\perp, k_\parallel)$ calculations revisit ICE observations from JET Pulse
No. 26148\cite{Cottrell1993,Dendy1995} and reproduce several important features:
the step change in signal strength from $\omega \leq 6\Omega_i$
to $\omega > 6\Omega_i$; the
doublet splitting of peaks at $\omega \approx 10\Omega_i$ in Fig.
\ref{fig:2D_F12_physicalpitch}, with origins visible in the fine $(k_\parallel,
k_\perp)$ structure in Fig. \ref{fig:2D_imag_physicalpitch}. The Doppler shift
for large $k_\parallel$ becomes of the order of the alpha-particle cyclotron
frequency making it difficult to determine which parts of the signal come
from which cyclotron harmonic numbers, which has important implications for
determining the major radius of origin of ICE.
Further effort is required to account for
the extra doublets visible in the continuous trace associated with JET Pulse No.
26148 in Fig. 2 of Ref. \cite{Cottrell1993}.

These calculations capture the finite perpendicular thermal spread of the alpha
particle distribution function, the integral of which has no known analytically
tractable form. The computational capabilities of the software,
LinearMaxwellVlasov.jl, used to generate the results further enable ICE as
a diagnostic for fusion plasmas since the code is capable of: calculating the
dispersion relation of plasmas consisting of an arbitrary number of species; and
of arbitrary distribution functions, although those with parallel Maxwellians,
like those presented here, are faster. It has been shown that it is possible to
qualitatively recreate signals containing lower harmonics of substantial amplitudes by
invoking 2D linear physics. The presence of ICE signals at lower harmonics in
PIC simulations has been attributed to nonlinear interactions between pairs of
more powerful higher harmonics\cite{Carbajal2014}. This work suggests that
observations of lower ICE harmonics may arise due to 2D
linear physics effects but does not exclude the possiblity that nonlinear 1D
physics\cite{Chapman2018} may also play a part.

>>>>

Observations of ion cyclotron emission (ICE) have been reported from
magnetically confined plasmas first in TFTR \cite{Cauffman1995,Dendy1995}
and JET \cite{Cottrell1988,Cottrell1993,Dendy1995}, followed by other tokamaks
PDX \cite{Heidbrink1994}, JT-60-U \cite{Ichimura2008}, DIII-D \cite{Heidbrink2011},
ASDEX \cite{DInca2014}, KSTAR \cite{Chapman2017,Chapman2018},
ASDEX-Upgrade \cite{Ochoukov2018},
TUMAN-3M \cite{Askinazi2018}, both NSTX and NSTX-U \cite{Fredrickson2019},
and EAST \cite{Liu2020}. Stellarators \cite{Saito2013} and
FRCs \cite{Nicks2020} have also exhibited this phenomenon. Energetic ion
populations that give rise to ICE have their origins in fusion
reactions in DT plasmas \cite{Cottrell2000, Cauffman1995},
ICRH \cite{Cottrell2000}, and neutral beam injection \cite{Ochoukov2018}. ICE
arising from fusion $\alpha$-particles is a proposed route to further
the understanding of burning plasmas in ITER \cite{Dendy2015,McClements2015}.

The foremost explanation of the mechanism responsible for ICE is the
magnetoacoustic cyclotron instability (MCI) \cite{Belikov1976}, which arises from
the free energy present in energetic ions with inverted distributions in
velocity space. Theoretical understanding of ICE is based on analytical
solutions to the linearised Maxwell-Vlasov system of equations \cite{Stix} where
the energetic ion species is modelled as ring-beam for which $f(v_\parallel,
v_\perp) \propto \exp(-(v_\parallel-u_\parallel)^2/ v_{th}^2)\delta(v_\perp -
u_\bot)$ \cite{Dendy1994, McClements1996}. Here $u_\parallel$, $u_\perp$ and
$v_{th}$ are the beam velocity (the signed drift speed along the magnetic
field), ring speed (the speed of the shifted centre of the inverted population
perpendicular to the magnetic field)
and thermal velocity, respectively. Subsequent studies augmented this
understanding via nonlinear fully-kinetic \cite{Cook2013,Chapman2017} and
hybrid \cite{Carbajal2014,Reman2019} particle-in-cell (PIC) simulations in one
spatial dimension with three vector components for velocities and electromagnetic
field (i.e. 1D3V), and recently 2D3V hybrid-PIC simulations \cite{Carbajal2021}. These
computational approaches operate under similar simplifications as the analytical
treatments in that the plasma is initially spatially homogeneous and periodic
in the spatial domain.
Linearised fluid approaches that incorporate toroidal geometry and
inhomogeneous plasma and magnetic fields have shed light of the eigenmode
structure of compressional Alfv{\'e}n eigenmodes, which as with ICE are also fast
Alfv{\'e}n modes driven by energetic ions, in spherical
tokamaks \cite{Sharapov2014, Gorelenkov2016} and conventional aspect ratio
tokamaks \cite{Kolesnichenko2000, Fulop2000, Heidbrink2006}.

The effectiveness of linear theory to reproduce the observed cyclotron
harmonic peaks in the fluctuations of the magnetic field was first explained
somewhat by 1D PIC simulations \cite{Cook2013}.
In these simulations the onset of nonlinearity
appeared to imprint the amplitudes of each of the harmonics into the signal as
they are at the end of the linear phase; each mode grows independently until 
growth is terminated for all modes when the fastest mode enters its nonlinear
phase meaning that both the amount of free energy available for instability has
decreased and that the unstable ion distribution function has been
perturbed into a more stable state. Analysis shows that only higher cyclotron
harmonics are destabilised \cite{McClements2015} to the MCI under these
particular conditions, however experimental
observations clearly indicate power at theoretically stable lower
harmonics \cite{Cottrell1988}. In a further step towards accurate recreation of
experimental observations, nonlinear phases of hybrid \cite{Carbajal2014} and
fully kinetic \cite{Cook2013,Chapman2017} PIC simulations went on to show that
wave-wave coupling was responsible for raising the amplitude of lower cyclotron
harmonics. Most recently, hybrid 2D simulations have begun to shed light on the
2D3V dynamics of ICE \cite{Carbajal2021}.

Finer detailed structure has been obvserved in ICE spectra, such as doublet
splitting of individual harmonics on JET \cite{Cottrell1993} and
DIII-D \cite{Thome2018} and even multiplet splitting on TUMAN-3M \cite{Askinazi2018},
ASDEX-Upgrade \cite{Ochoukov2019} and EAST \cite{Liu2020}, which has been made
possible by advances in resolution of diagnostic techniques.
This paper highlights one underlying kinetic mechinism that produces
doublet splitting due to Doppler shifted resonances and the
intricate landscape of the MCI growth rate on the fast Alfv{\'e}n dispersion
surface in $(k_\bot, k_\parallel)$ space.
In principle, splitting could be produced by grad-B or curvature
drifts \cite{Cottrell1993,Fulop1998} or potentially any mechanism that is capable
of adding a Doppler shift to the wave-particle resonance condition. Note that
this study focusses on edge-ICE, named for the region of plasma from which the signal
arises, as opposed to core-ICE \cite{Ochoukov2018, Liu2020}.

To better enable ICE as a diagnostic for confined fast ions, it is necessary to
further develop our understanding with a range of tools. This work presents
the growth rates in 2D wavenumber space of the MCI under the conditions of
outer mid-plane of the JET PTE \cite{Cottrell1993} Pulse No. 26148. These growth
rates are obtained from a dispersion relation solver,
LinearMaxwellVlasov.jl \cite{LinearMaxwellVlasov.jl}, written in the Julia programming
language \cite{Bezanson2017}. These results narrow the gap between analytical
theory and expensive HPC simulations and can be run on desktop computers
in a matter of minutes, further enabling research into this topic.

In section 2, the computational method is described for calculating
the linear growth rates of the MCI, which are presented for energetic ring-beam
alpha-particles with non-zero perpendicular thermal spread in section 3
where a comparison is made to the observations on JET.
Results are summarised in section 4.

@@

Solutions to the linearised Maxwell-Vlasov system of equations are found in
perpendicular and parallel wavenumber space, $(k_\bot, k_\parallel)$, for a
ring-beam distribution of energetic minority
alpha-particles, Maxwellian electrons and background thermal majority
deuteron ions. The ring-beam distribution function, Eq. 1, is a separable
function of the beam in $v_\parallel$ given by Eq. 2 and the
ring in $v_\perp$ given by Eq. 3.
In the context of the fully kinetic Maxwell-Vlasov linear theory,
which rests upon the assumption of
an infinite homogeneous plasma and a straight uniform magnetic field,
this distribution function is parameterised by the drift
velocity parallel to the magnetic field $u_\parallel$, the speed of the
Maxwellian distribution perpendicular to the magnetic
field $u_{\perp}$, and the isotropric thermal velocity $v_{th}$. The
thermal spread of the alpha-particles' distribution function 
$v_{th}=v_{\parallel,th}=v_{\perp,th}=u_0/100$. The number density
ratio of alpha-particles to the majority fuel deuterons is $1.5\times10^{-4}$.

@@

Although curvature and grad-B drift terms would be present in fast
alpha-particle populations on JET and these drifts would alter the wave-particle
resonance condition, it is worth noting that they are both of
order of the ratio of the Larmor radius to the machine size, $\rho^*$,
and have been neglected in these calculations where $u_\bot \gg
\rho^* u_0$ is always the case.

Solutions to the linearised Maxwell-Vlasov system of equations are given by the
dispersion relation, Eq. \ref{eq:dettensor}, which is satisfied for sets of
$(\omega,\vec{k})$ 
along the fast Alfv{\'e}n branch  on a grid of $(k_\parallel, k_\bot)$.
The hot dielectric tensor \cite{Stix} $\vec{\vec{\epsilon}}$, given by Eq.
\ref{eq:dielectrictensor}, is calculated for each species, $s$,

@@

Integrals over perpendicular velocity required for calculating the
dielectric tensor for Maxwellian distribution functions are given in terms of
modified Bessel functions of the first kind \cite{Stix} for both electrons and
deuterons i.e. when $u_\perp =0$. The same perpendicular integrals over the
alpha-particles' ring distribution given by Eq. 3 are calculated numerically via
Gauss-Kronrod \cite{Kronrod1965} quadrature using the GaussGK.jl \cite{QuadGK}
library. An absolute and relative tolerance of $0$ and $10^{-8}$, respectively,
is used to confirm convergence of the integral values.
Since all the parallel distribution functions for all species are Maxwellians,
the parallel integrals can be calculated from the well known plasma dispersion
function\cite{Fried1961}, which is laid out in the appendix.

Bessel function indices, $n$, in Eq. \ref{eq:dielectrictensor} are incremented
outwards from 0 in both the positive and negative directions
until the relative change between the current summand and that of the
previous iteration is less than a tolerance of $10^{-8}$. 
Root finding of Eq. \ref{eq:dettensor} in the complex frequency plane
for a given $(k_\bot, k_\parallel)$-pair is performed by an amalgamation of the
Nelder-Mead \cite{Nelder1965} optimisation method with the winding number method.
If the complex values at the nodes of the Nelder-Mead simplex, which in this
case is just a triangle, have a winding number
of zero and therefore the simplex does not contain a root then the optimisation
procedure follows the Nelder-Mead method, otherwise the simplex bifurcates.
The child simplex that contains the root is used for the next
iteration. The optimisation iteration procedure finishes successfully when
the simplex that encircles a root is small enough such that its extent in real
(imaginary) frequency is less than $10^{-4}\, \Omega_i$ $(10^{-5}\, \Omega_i)$,
where $\Omega_i$ is the cyclotron frequency of both
the alpha-particles and deuterons since they share a charge-mass ratio.
This approach is found to be adequately robust and
faster than either the Nelder-Mead method or winding number approach used
separately. Each 512 by 512 scan in $(k_\perp, k_\parallel)$ takes between 5
and 6 minutes on a current laptop.

@@

The growth rates of unstable modes for the case where the pitch-angle cosine
value is zero ($u_\parallel = 0$ in Eq. 1) are plotted in Fig.
\ref{fig:2D_F12_zeropitch}(a) against the associated real frequency (only modes
with growth rates in excess of the solver tolerance are shown i.e. $\Im(\omega)
> 10^{-5} \Omega_i$). In this panel shading indicates parallel wavenumber in
units of $\Omega_i/V_A$. Fig. \ref{fig:2D_F12_zeropitch}(b) plots the same data
in the $(\omega,k_\parallel)$ plane where shading instead indicates growth
rate in units of $\Omega_i$.
The simplicity of the consecutive unstable cyclotron harmonics visible in
panel (a) betrays the complexity of the structure in the $(\omega,k_\parallel)$
plane. Integration over $k_\parallel$ projects the solutions onto the frequency
axis, which loses the structure and produces narrow harmonics. Figure
\ref{fig:2D_imag_zeropitch} gives a $(k_\perp, k_\parallel)$ space view of the
same data as Fig. \ref{fig:2D_F12_zeropitch} as well as damped and
normal modes, except shading indicates growth rate. Overlaid
contours indicate the real frequency in units of ion cyclotron harmonics. Only
narrow regions around the cyclotron harmonics are unstable. Overlaid straight
traces show propagation angle. No Doppler shift is present due to the lack
of parallel drift in the alpha-particle ring-beam. The growth rate increases
approximately linearly with cyclotron harmonic number, which is consistent with
the literature \cite{Dendy1994}. Together these figures serve to introduce a 2D
view (growth rates in $(k_\perp, k_\parallel)$ space) of traditionally 1D data
(growth rates against frequency). The importance of a non-zero $u_\parallel$ is
discussed next.

@@

Fig. \ref{fig:2D_F12_physicalpitch} plots solutions for the physically more
realistic pitch case ($u_\parallel = -0.646 u_0$) in the same manner as Fig.
\ref{fig:2D_F12_zeropitch} except there is an additional black trace in the
upper panel, which represents the measured JET ICE intensity signal; the continuous
trace from Fig. 2 in Ref. \cite{Cottrell1993}.
These observations of ICE intensity are plotted on a log scale scaled to match
the range of the linear MCI growth rates.
Direct comparison of log-scale
ICE observations and linear growth rates is justified to the extent that
nonlinear PIC simulations indicate that the linearly unstable modes rising
as $\exp(\gamma t)$, where $\gamma$ is a mode's growth rate,
cease to grow once the fastest growing mode becomes
nonlinear \cite{Cook2013}.
Therefore under the assumption that all modes start growing in unison from
identical seed amplitudes then the log of a mode's amplitude is
proportional to its growth rate.

@@

Doublet splitting can be seen in Fig. \ref{fig:2D_imag_physicalpitch}, which is the
counterpart figure to Fig. \ref{fig:2D_F12_physicalpitch} as Fig.
\ref{fig:2D_imag_zeropitch} is to Fig. \ref{fig:2D_F12_zeropitch},
at e.g. the $12^{th}$ harmonic  near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$, where $V_A$ is the Alfv{\'e}n speed.
Qualitatively, scattered points in Fig.
\ref{fig:2D_F12_physicalpitch}(a) resembles the experimental observations given
by the black trace in that the
signal below the $7^{th}$ cyclotron harmonics is lower and flatter with well
defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change
larger and the harmonics are less easily identifiable. These features cannot be
be recreated with a 1D calculation for these distributions functions
since they arise from contributions from a range of propagation angles.
Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes
more clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is
characterised by narrow regions of instability near Doppler shifted cyclotron
harmonics that are stable when $k_\parallel=0$ and unstable at larger
negative parallel wavenumbers. The higher frequency behaviour in the region
$k_\bot \geq 7\Omega_i/V_A$ is dominated by broader bands of instability
near to Doppler shifted cyclotron harmonics, which have
an intricate structure in wavenumber space. Modes numbering 1-5 and
14 are stable when propagating exactly perpendicular whilst obliquely
propagating modes remain unstable. The magnitude of the Doppler shift of
the growing modes can reach $\approx \Omega_i$ for large propagation angles at 
lower frequencies and also for
oblique angles at higher frequencies; see, for example, the mode
associated with the $15^{th}$ cyclotron harmonic near $k_\parallel \simeq
-\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 13 \Omega_i/V_A$, i.e. a propagation
angle of $\sim 93^\circ$. Furthermore, the breadth of the harmonics can be
attributed to the contributions from multiple propagation angles even though the
lower harmonics are largely stable for forward travelling modes and
propagation angles that are close to perpendicular.

Figure \ref{fig:2D_real_zeropitch}
shows the smoothed real frequency part of the solution, as depicted
by contours on Figs. \ref{fig:2D_imag_zeropitch} and
\ref{fig:2D_imag_physicalpitch}, for the case of
zero pitch angle cosine and is normalised by
the high frequency expression for the fast Alfv{\'e}n wave set out in Eq.
25 of Ref. \cite{Dendy1994} (also Eq. 11 of Ref. \cite{McClements1996}),

@@

\noindent where $k^2 = k_\parallel^2 + k_\perp^2$. There is no discernable
difference for the non-zero pitch case.

@@

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig4.pdf}
    \caption{(colour online) As in Fig. \ref{fig:2D_imag_zeropitch}, except for
    a ring-beam of alpha-particles with a more realistic pitch-angle value
    $u_{\parallel} = -0.646 u_0 \simeq -1.1 V_A$. The non-zero parallel drift
    speed causes up-down asymmetry; this figure for a pitch-angle cosine of the
    opposite sign, $u_{\parallel} = 0.646 u_0$, is identical except flipped
    about $k_\parallel=0$.}
    \label{fig:2D_imag_physicalpitch}
\end{figure}

@@

This paper presents the linear Maxwell-Vlasov dispersion relation of a plasma
consisting of thermal electrons, thermal deuteron fuel ions and a minority
energetic ring-beam of fusion born alpha-particles. These high resolution
$(k_\perp, k_\parallel)$ calculations revisit ICE observations from JET Pulse
No. 26148 \cite{Cottrell1993,Dendy1995} and reproduce several important features:
the step-change in signal strength from $\omega \leq 7\Omega_i$
to $\omega > 7\Omega_i$; the
doublet splitting of peaks at $\omega \approx 10\Omega_i$ in Fig.
\ref{fig:2D_F12_physicalpitch}, with origins visible in the fine $(k_\bot,
k_\parallel)$ structure in Fig. \ref{fig:2D_imag_physicalpitch}. The Doppler shift
for large $k_\parallel$ becomes of the order of the alpha-particle cyclotron
frequency making it difficult to determine which parts of the signal come
from which cyclotron harmonic numbers, which has important implications for
determining the major radius of origin of ICE.
Further effort is required to account for
the extra doublets visible in the continuous trace associated with JET Pulse No.
26148 in Fig. 2 of Ref. \cite{Cottrell1993}.

These calculations capture the finite perpendicular thermal spread of the alpha
particle distribution function, the integral of which has no known analytically
tractable form. The computational capabilities of the software,
LinearMaxwellVlasov.jl, used to generate the results further enable ICE as
a diagnostic for fusion plasmas since the code is capable of calculating the
dispersion relation of plasmas consisting of an arbitrary number of species and
of arbitrary distribution functions, although those with parallel Maxwellians,
like those presented here, are handled faster.
It has been shown that it is possible to
qualitatively recreate signals containing lower harmonics of substantial amplitudes by
invoking 2D linear physics. The presence of ICE signals at lower harmonics in
PIC simulations has been attributed to nonlinear interactions between pairs of
more powerful higher harmonics \cite{Carbajal2014}. This work suggests that
observations of lower ICE harmonics may arise due to 2D
linear physics effects but does not exclude the possiblity that nonlinear 1D
physics \cite{Chapman2018} may also play a part.

There are several important features missing from these calculations. Among them
are the effects of the spatial inhomogeniety of the magnetic field and background
plasma giving rise to drifts,
and the quantisation arising from toroidal periodicity. These
effects may create an eigenmode structure, not accounted for herein,
associated with the kinetic MCI.
Cottrell et al. \cite{Cottrell1993} discusses the implications of propagation
angles on ICE signals and indicates that lower harmonics must propagate
increasingly close to exactly perpendicular. It remains the subject of
further work to incorporate these effects into linear and nonlinear studies of ICE.
